A Triple Origin for Twin Blue Stragglers in Compact Binaries

### 2019-01-04 BSS7782.tex
fixed references to be in Simon's format

Blue straggler (BS) stars are brighter and bluer than the
main-sequence turn-off (MSTO) in a cluster colour-magnitude diagram
\citep[e.g.][]{1953AJ.....58...61S,simunovic14,simunovic16}.  Two
primary channels for BS formation have been proposed: mass transfer
from an evolved donor on to a main-sequence star in a binary star
system
\citep[e.g.][]{mccrea64,1997A&A...328..143P,knigge09,mathieu09,leigh11,geller11,geller12,gosnell14,gosnell15},
and direct stellar collisions involving main-sequence stars possibly
mediated via binaries
\citep[e.g.][]{hills75,1997A&A...328..130P,shara97,leigh07,leigh11,leigh13,hypki13,2018arXiv181100058P}.
Other possible, albeit related, formation mechanism includes mergers
of close main-sequence binaries (MS-MS) \cite{2018arXiv181100058P},
and mergers of the inner binaries of hierarchical triple star systems
induced by Lidov-Kozai oscillations coupled with tidal damping
\citep[e.g.][]{perets09}.

In spite of these specific predictions for the expected properties of
BSs formed from each of the above production mechanisms, there exist
many BSs with observed properties that defy these simple scenarios.
For example, in the old open cluster M67, there lurks a candidate
triple system that is posited to host two BSs
\citep{vandenberg01,sandquist03}.  The observations suggest that the
outer tertiary is itself a BS, with a mass $\sim$ 1.7 M$_{\odot}$ and
orbiting the inner binary with a period of 1188.5 days
\citep{sandquist03}.  The inner binary has a period of only 1.068 days
\citep{vandenberg01}, and hosts a BS of mass 2.52 M$_{\odot}$.  In
order to form this system \del{after convolving its observed
  properties with its host cluster age,} we require at least five
stars in order to conserve the total system mass \citep{leigh11}.  As
explained in \citet{leigh11}, this is strongly indicative of a
dynamical origin for the system, and a single direct interaction
involving a binary and a triple that resulted in two separate
collision events is the most probable explanation for its origin
(i.e., a single interaction involving two multiples with two or more
stars is a more likely scenario to produce this system than two
back-to-back direct binary-binary interactions)
\citep{2004MNRAS.350..615G}.  \simon{Maybe add something about the
  curious binary AE Aurigae? - NL2: For sure.  I still don't know anything about it... But if I get time over the next few days I will do my research and write something.}

Even more curious, there exists in the old open cluster NGC 188 a
double BS binary, called Binary 7782.  More generally, the BS
population in NGC 188 has a bi-modal period-eccentricity distribution.
As discussed in \citet{leigh11}, this could be hinting at a triple
origin for at least some subset of the total BS population.  As for
Binary 7782, \citet{mathieu09} observed a compact and mildly eccentric
(i.e., $e \sim 0.1$) binary star system with an orbital period of
$\sim$ 10 days hosting two roughly equal-mass blue stragglers.  During
a given binary-binary interaction, the probability that not one but
two direct (MS-MS) collisions will occur is less than $10^{-2}$
\citep{leonard89,leigh11,leigh12}.  Plus, binaries with collision products
typically have relatively long orbital periods
\cite{2011Sci...334.1380F}. Dynamically, it is difficult to form a
short-period binary composed of two collision products during a
collisional interaction in a star cluster \citep{2011Sci...334.1380F}.
So, how did Binary 7782 form?

@@

We consider a scenario where the outer star is filling its Roche lobe
and transfers mass to the inner binary.  The mass transfer stream
gathers at the circularization radius a$_{\rm c}$, and forms a circumbinary disk
around the inner binary.  Using conservation of angular momentum, we
equate the specific angular momentum of the accreted mass at the inner
Lagrangian point of the (outer) donor star to the final specific
angular momentum of the accretion stream at the circularization radius
about the inner binary, this results in
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - R_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer
tertiary companion, $a_{\rm c}$ is the semi-major axis of the orbit
about the inner binary corresponding to the circularization radius and
v$_{\rm orb,c}$ is the orbital velocity at $a_{\rm c}$.  The distance
from the center of mass corresponding to the outer tertiary companion
defined by the Roche lobe is given by Equation 2 in
\citep{eggleton83}.  Combining Equation 2 in \citet{eggleton83} (with
mass ratio q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$m$_{\rm 2}$)) with
Equation~\ref{eqn:specangmom1}, we can solve for the circularization
radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we
require that a$_{\rm in} <$ a$_{\rm c}$.

@@

\begin{figure}[ht!]
\plotone{fig2.eps}
\caption{Parameter space in the P$_{\rm out}$-P$_{\rm in}$-plane
  allowed for the hypothetical outer tertiary orbit of Binary 7782.
  The solid diagonal black line shows the period corresponding to the
  circularization radius a$_{\rm c}$ for the mass transfer stream
  coming from the outer star (i.e., at the onset of mass transfer), assuming a mass ratio of q $=$ 0.7 and component masses 
  of m$_{\rm 1} =$ 0.9 M$_{\odot}$, m$_{\rm 2} =$ 1.1 M$_{\odot}$ and 
  m$_{\rm 3} =$ 1.4 M$_{\odot}$. \simon{what is the purpose of extending q to 0.1 and
    10? - NL2: I was originally thinking that q changes during MT, but I have now removed the other lines and just left the one corresponding to our test case.} We assume initial component masses of m$_{\rm 1}$ = 1.1
  M$_{\rm \odot}$ and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner
  binary components, and m$_{\rm 3}$ is computed for the outer
  tertiary according to our assumed mass ratio (with our fiducial case
  corresponding to q $=$ 0.7).  We assume completely conservative mass
  transfer for this exercise, and a final mass for the outer tertiary
  of 0.6 M$_{\rm \odot}$ once it has become a WD.  The dashed diagonal
  black line shows a rough criterion for dynamical stability in the
  triple, approximately following \citet{mardling99} (i.e., a$_{\rm
    in} <$ 0.1a$_{\rm out}$ is required for long-term dynamical
  stability in equal-mass co-planar triples).  The vertical solid red
  lines show the outer orbital periods corresponding to the hard-soft
  boundary assuming central velocity dispersions of $\sigma =$ 1, 5
  and 10 km s$^{-1}$.  The vertical dashed black lines show the
  maximum outer orbital period P$_{\rm out}$ for which the outer
  tertiary companion is Roche lobe-filling, assuming a stellar radius
  of R$_{\rm 3} =$ 200 R$_{\odot}$.  The horizontal dashed red line shows the observed
  orbital period for Binary 7782, using its observed orbital period
  and our assumed final inner companion masses (i.e., m$_{\rm 1} =$
  m$_{\rm 2} =$ 1.4 M$_{\odot}$).  Finally, the thick solid horizontal
  red line shows the parameter space for P$_{\rm out}$ allowed after
  considering all of the aforementioned criteria.  \simon{I think that
    the figure needs work. I like putting the parameter space in terms
    of orbital periods, but the q=0.1 and q=10 constraints seems
    not very practical. The red labels are not readable, and I am not
    really sure how to read the figure. What about the alternative in
    the next figure, which shows tertiary mass vs. orbital period? - NL2: I have now edited this figure, removed the other q curves, increased font size, etc., but am okay with it if you want to remove it for something else.}
\label{fig:fig2}}
\end{figure}

@@

With these parameters the orbital separation of the outer binary
becomes $\sim 3.86$\,au, and we adopt the outer orbit to be circular.
We choose a relative initial inclination between the inner and the
outer orbits of $i = 0^\circ$, $i = 9^\circ$, $i = 90^\circ$ and $i =
180^\circ$.  The first two choices are motivated by observed low-mass stellar
triples composed of low-mass stars which have been observed to favor
co-planar configurations \citep{moe18}. To test the effect of highly
inclined and retrogade orbits we include the $i = 90^\circ$ and $i =
180^\circ$ cases.  In Fig.\,\ref{fig:topview_at_t0} we present a top
view of the triple system in our simulations at the onset of Roche-lobe overflow.

@@

\item The BSs in the inner binary should have roughly equal masses,
  independent of their initial masses.  This is because it is the
  lowest mass object that typically accretes the fastest, since its
  orbital velocity and distance relative to the circumbinary disk is
  typically the lowest
  \citep[e.g.][]{kroupa95,bate97,bate00,haiman09,bate12,farris15,rafikov16,kelley17}.
  This quickly brings the mass ratio toward unity.  This predicts that
  the initially lower mass MS star should accrete the most, and should
  hence be enriched more by any accreted material.  This could be
  observable in the surface layers of a radiative star.  If the donor
  is an RGB star, the accretor will be enriched in mostly carbon,
  oxygen and helium.  If the donor is an AGB star, it will be enriched
  in mostly s-process elements.

\item Twin BSs in compact binaries formed from the mechanism proposed here 
  should be more frequent in younger clusters with ages $\lesssim$ 4-6
  Gyr.  This is because clusters with a main-sequence turn-off mass
  $\lesssim$ 1.2 M$_{\odot}$ have convective envelopes
  \citep[e.g.][]{iben91,maedoer09}, and a radiative envelope for the
  donor in a mass transferring binary ensures stable accretion on to
  the accretor.  \simon{Can we quantify this?}

@@

We emphasize in closing that our choice for the initial mass of the outer tertiary is critical since it
ensures that the donor star during the mass transfer process will have
a radiative envelope \citep[e.g.][]{maeder09}.  In turn, this ensures
that the mass transfer will be maximally conservative, such that the
accretion stream will be maximally stable, accreting at a stable and
roughly constant rate \citep[e.g.][]{iben91}.

>>>>

Blue straggler (BS) stars are brighter and bluer than the
main-sequence turn-off (MSTO) in a cluster colour-magnitude diagram
\citep[e.g.][]{1953AJ.....58...61S,2014ApJ...782...49S}.  Two
primary channels for BS formation have been proposed: mass transfer
from an evolved donor on to a main-sequence star in a binary star
system
\citep[e.g.][]{1964MNRAS.128..147M,1997A&A...328..143P,2009Natur.457..288K,2011MNRAS.410.2370L,2011Natur.478..356G},
and direct stellar collisions involving main-sequence stars possibly
mediated via binaries
\citep[e.g.][]{hills75,1997A&A...328..130P,2007ApJ...661..210L,2013MNRAS.428..897L,2013MNRAS.429.1221H,2018arXiv181100058P}.
Other possible, albeit related, formation mechanism includes mergers
of close main-sequence binaries (MS-MS) \cite{2018arXiv181100058P},
and mergers of the inner binaries of hierarchical triple star systems
induced by Lidov-Kozai oscillations coupled with tidal damping
\citep[e.g.][]{2009ApJ...697.1048P}.

In spite of these specific predictions for the expected properties of
BSs formed from each of the above production mechanisms, there exist
many BSs with observed properties that defy these simple scenarios.
For example, in the old open cluster M67, there lurks a candidate
triple system that is posited to host two BSs
\citep{2001A&A...375..375V,2003AJ....125..810S}.  The observations suggest that the
outer tertiary is itself a BS, with a mass $\sim$ 1.7 M$_{\odot}$ and
orbiting the inner binary with a period of 1188.5 days
\citep{{2003AJ....125..810S}.  The inner binary has a period of only 1.068 days
\citep{2001A&A...375..375V}, and hosts a BS of mass 2.52 M$_{\odot}$.  In
order to form this system \del{after convolving its observed
  properties with its host cluster age,} we require at least five
stars in order to conserve the total system mass \citep{2011MNRAS.410.2370L}.  As
explained in \citet{2011MNRAS.410.2370L}, this is strongly indicative of a
dynamical origin for the system, and a single direct interaction
involving a binary and a triple that resulted in two separate
collision events is the most probable explanation for its origin
(i.e., a single interaction involving two multiples with two or more
stars is a more likely scenario to produce this system than two
back-to-back direct binary-binary interactions)
\citep{2004MNRAS.350..615G}.  \simon{Maybe add something about the
  curious binary AE Aurigae? - NL2: For sure.  I still don't know anything about it... But if I get time over the next few days I will do my research and write something.}

Even more curious, there exists in the old open cluster NGC 188 a
double BS binary, called Binary 7782.  More generally, the BS
population in NGC 188 has a bi-modal period-eccentricity distribution.
As discussed in \citet{2011MNRAS.410.2370L}, this could be hinting at a triple
origin for at least some subset of the total BS population.  As for
Binary 7782, \citet{2009Natur.462.1032M} observed a compact and mildly eccentric
(i.e., $e \sim 0.1$) binary star system with an orbital period of
$\sim$ 10 days hosting two roughly equal-mass blue stragglers.  During
a given binary-binary interaction, the probability that not one but
two direct (MS-MS) collisions will occur is less than $10^{-2}$
\citep{1989AJ.....98..217L,2011MNRAS.410.2370L,2012MNRAS.425.2369L}.  Plus, binaries with collision products
typically have relatively long orbital periods
\cite{2011Sci...334.1380F}. Dynamically, it is difficult to form a
short-period binary composed of two collision products during a
collisional interaction in a star cluster \citep{2011Sci...334.1380F}.
So, how did Binary 7782 form?

@@

We consider a scenario where the outer star is filling its Roche lobe
and transfers mass to the inner binary.  The mass transfer stream
gathers at the circularization radius a$_{\rm c}$, and forms a circumbinary disk
around the inner binary.  Using conservation of angular momentum, we
equate the specific angular momentum of the accreted mass at the inner
Lagrangian point of the (outer) donor star to the final specific
angular momentum of the accretion stream at the circularization radius
about the inner binary, this results in
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - R_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer
tertiary companion, $a_{\rm c}$ is the semi-major axis of the orbit
about the inner binary corresponding to the circularization radius and
v$_{\rm orb,c}$ is the orbital velocity at $a_{\rm c}$.  The distance
from the center of mass corresponding to the outer tertiary companion
defined by the Roche lobe is given by Equation 2 in
\citep{1983ApJ...268..368E}.  Combining Equation 2 in \citet{1983ApJ...268..368E} (with
mass ratio q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$m$_{\rm 2}$)) with
Equation~\ref{eqn:specangmom1}, we can solve for the circularization
radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we
require that a$_{\rm in} <$ a$_{\rm c}$.

@@

\begin{figure}[ht!]
\plotone{fig2.eps}
\caption{Parameter space in the P$_{\rm out}$-P$_{\rm in}$-plane
  allowed for the hypothetical outer tertiary orbit of Binary 7782.
  The solid diagonal black line shows the period corresponding to the
  circularization radius a$_{\rm c}$ for the mass transfer stream
  coming from the outer star (i.e., at the onset of mass transfer), assuming a mass ratio of q $=$ 0.7 and component masses 
  of m$_{\rm 1} =$ 0.9 M$_{\odot}$, m$_{\rm 2} =$ 1.1 M$_{\odot}$ and 
  m$_{\rm 3} =$ 1.4 M$_{\odot}$. \simon{what is the purpose of extending q to 0.1 and
    10? - NL2: I was originally thinking that q changes during MT, but I have now removed the other lines and just left the one corresponding to our test case.} We assume initial component masses of m$_{\rm 1}$ = 1.1
  M$_{\rm \odot}$ and m$_{\rm 2}$ = 0.9 M$_{\rm \odot}$ for the inner
  binary components, and m$_{\rm 3}$ is computed for the outer
  tertiary according to our assumed mass ratio (with our fiducial case
  corresponding to q $=$ 0.7).  We assume completely conservative mass
  transfer for this exercise, and a final mass for the outer tertiary
  of 0.6 M$_{\rm \odot}$ once it has become a WD.  The dashed diagonal
  black line shows a rough criterion for dynamical stability in the
  triple, approximately following \citet{mardling99} (i.e., a$_{\rm
    in} <$ 0.1a$_{\rm out}$ is required for long-term dynamical
  stability in equal-mass co-planar triples).  The vertical solid red
  lines show the outer orbital periods corresponding to the hard-soft
  boundary assuming central velocity dispersions of $\sigma =$ 1, 5
  and 10 km s$^{-1}$.  The vertical dashed black lines show the
  maximum outer orbital period P$_{\rm out}$ for which the outer
  tertiary companion is Roche lobe-filling, assuming a stellar radius
  of R$_{\rm 3} =$ 200 R$_{\odot}$ (which corresponds to the maximum stellar radius reached on the AGB for the range of tertiary masses of interest to us.  The horizontal dashed red line shows the observed
  orbital period for Binary 7782, using its observed orbital period
  and our assumed final inner companion masses (i.e., m$_{\rm 1} =$
  m$_{\rm 2} =$ 1.4 M$_{\odot}$).  Finally, the thick solid horizontal
  red line shows the parameter space for P$_{\rm out}$ allowed after
  considering all of the aforementioned criteria.  \simon{I think that
    the figure needs work. I like putting the parameter space in terms
    of orbital periods, but the q=0.1 and q=10 constraints seems
    not very practical. The red labels are not readable, and I am not
    really sure how to read the figure. What about the alternative in
    the next figure, which shows tertiary mass vs. orbital period? - NL2: I have now edited this figure, removed the other q curves, increased font size, etc., but am okay with it if you want to remove it for something else.}
\label{fig:fig2}}
\end{figure}

@@

With these parameters the orbital separation of the outer binary
becomes $\sim 3.86$\,au, and we adopt the outer orbit to be circular.
We choose a relative initial inclination between the inner and the
outer orbits of $i = 0^\circ$, $i = 9^\circ$, $i = 90^\circ$ and $i =
180^\circ$.  The first two choices are motivated by observed low-mass stellar
triples composed of low-mass stars which have been observed to favor
co-planar configurations \citep{2018ApJ...854...44M}. To test the effect of highly
inclined and retrogade orbits we include the $i = 90^\circ$ and $i =
180^\circ$ cases.  In Fig.\,\ref{fig:topview_at_t0} we present a top
view of the triple system in our simulations at the onset of Roche-lobe overflow.

@@

\item The BSs in the inner binary should have roughly equal masses,
  independent of their initial masses.  This is because it is the
  lowest mass object that typically accretes the fastest, since its
  orbital velocity and distance relative to the circumbinary disk is
  typically the lowest
  \citep[e.g.][]{1995MNRAS.277.1491K,1997MNRAS.285...33B,2000MNRAS.314...33B}.
  This quickly brings the mass ratio toward unity.  This predicts that
  the initially lower mass MS star should accrete the most, and should
  hence be enriched more by any accreted material.  This could be
  observable in the surface layers of a radiative star.  If the donor
  is an RGB star, the accretor will be enriched in mostly carbon,
  oxygen and helium.  If the donor is an AGB star, it will be enriched
  in mostly s-process elements.

\item Twin BSs in compact binaries formed from the mechanism proposed here 
  should be more frequent in younger clusters with ages $\lesssim$ 4-6
  Gyr.  This is because clusters with a main-sequence turn-off mass
  $\lesssim$ 1.2 M$_{\odot}$ have convective envelopes
  \citep[e.g.][]{1991ApJS...76...55I,2009pfer.book.....M}, and a radiative envelope for the
  donor in a mass transferring binary ensures stable accretion on to
  the accretor.  \simon{Can we quantify this?}

@@

We emphasize in closing that our choice for the initial mass of the outer tertiary is critical since it
ensures that the donor star during the mass transfer process will have
a radiative envelope \citep[e.g.][]{2009pfer.book.....M}.  In turn, this ensures
that the mass transfer will be maximally conservative, such that the
accretion stream will be maximally stable, accreting at a stable and
roughly constant rate \citep[e.g.][]{1991ApJS...76...55I}.
